---
title: "Example Vignette: R Package fmatrix"
author: 'R Samyak'
date: "29/07/2020"
output: html_document
---

We first load the R package:
```{r setup, include = TRUE}
library(fmatrix)
```


# Introduction

An $F$-matrix is a lower triangular non-negative integer matrix of a particular form, introduced in Kim et al. (2020).

An example of an $F$-matrix:
```{r}
## Note: this function is only for illustrative purposes,
## does not generate an $F$-matrix from any distribution (yet).
set.seed(8964)
mat <- rFmat(8)

mat
```


## Check if a given matrix is an $F$-matrix

We can check if a given object is an $F$-matrix or not.  Note that this check is expensive, and must be used sparingly for large $n$.
```{r}
is.Fmat(mat)
```


```{r}
not_mat <- mat
not_mat[7, 1] <- 0
not_mat[5, 3] <- 0

not_mat

is.Fmat(not_mat)
```

For a given matrix that is not an $F$-matrix, we can also see where the checks fail:
```{r}
where.Fmat.debug(not_mat)
```


## Pre-loaded lists of $F$-matrices
The package contains pre-computed lists of $F$-matrices for $n = 5, ..., 11$ as well as $l^1$ and $l^2$ distances for $n = 5, ..., 9$.

```{r}
mat2 <- F.list8[[5]]
```

Note there are $5$ $F$-matrices corresponding to $n=5$.  The order is determined as in $F.list5$, which is available in the package.
```{r}
Dmat.l2.5
```

## Distances between $F$-matrices

We can compute the $l^1$ and $l^2$ distances between two given $F$-matrices.
```{r}
distance_Fmat(mat, mat2, dist = "l2")
```

## $F$-matrices with coalescent times

If we have the branch lengths as well:
```{r}
n <- 8

times <- rev(sort(sample(1:50, n-1)))
times2 <- rev(sort(sample(1:50, n-1)))

wt <- wt_from_times(times)
wt2 <- wt_from_times(times2)

wFmat <- list(f = mat,
              w = wt)

wFmat2 <- list(f = mat2,
               w = wt2)

fmatrix:::distance_Fmat_weighted(wFmat, wFmat2, dist = c("l1"))
```


We can convert a given weight matrix back to coalescent times as well:
```{r}
times_from_wt(wt)
```

## Constructing full list of $F$-matrices

We can use the the `construct_trees` function, borrowed from `JuliaPalacios/phylodyn` to explicitly compute all $F$-matrices for a given $n$.  This should not be used for $n > 11$, since it will take literally forever to run.
```{r}
n <- 8
res<-construct_trees(n)
nrow(res$res)

F.list<-list()
for (j in 1:nrow(res$res)){
  F.list[[j]]<- Fmat_from_construct(n, res$res[j,], res$F.matrixRel)
}

n.tr <- length(F.list)
```
Indeed, there are 272 $F$-matrices for $n = 8$.


# Models and Likelihoods

## Kingman model

For a given $F$-matrix, we can calculate its likelihood under the Kingman model:
```{r}
kingman_likelihood(mat)
```

## Blum-Francois $\beta$-splitting model

The Blum-Francois $\beta$-splitting model contains the special case of the Kingman model, which corresponds to $\beta = 0$.  We can calculate the likelihood of a given $F$-matrix under the BF model for different values of $\beta$ in $[-1, \infty)$.
```{r}
beta_BF_likelihood(mat, betas = c(0, 4, -1, -2,  1e7))
```




# Summary statistics for $F$-matrices

## Explicit sample Frechet Mean and Variance

We can calculate the sample Frechet mean and variance for a given sample of $F$-matrices.

```{r}
sampleF <- sample(1:length(F.list8), 1000, replace = TRUE)
sampleF <- F.list8[sampleF]

fmatrix:::frechet_var_sample(sampleF, dist = "l1")
```

We can also calculate the population version under a given model if we know the likelihoods (such as under the BF $\beta$ model)

```{r}
fullF.list <- F.list8
probs <- sapply(F.list8, beta_BF_likelihood, betas = 1.5)

fmatrix:::frechet_var_pop(fullF.list, probs, dist = "l2")
```

For larger $n$, where explicit computation of Frechet mean is intractable, we include functions that allow for constructing a Markov chain on the space of $F$-matrices which can be used for combinatorial optimisation.  (Check `simulations_compare.Rmd`.)


# Miscellaneous

## Converting to and from `phylo` objects

```{r}
library(ape)
## TODO: This wraps around `tree_from_F` (Jaime's code).  There seems to
## be a bug in that for large n.
t <- mytree_from_F(mat2)

t

plot(t, direction = "downwards")
```

## Plotting trees corresponding to $F$-matrices

We can plot the tree directly from $F$-matrix form as well:
```{r}
plotF(mat2)
```

`plotF.list` is a simple wrapper around `plotF` for a list of $F$-matrices: 
```{r fig.width = 6, fig.height = 6}
fmatrix:::plotF.list(F.list6)
```


## Converting to and from $D$-matrices

$D$-matrices are used in proving the bijection between the space of $F$-matrices and the spaes of ranked oriented trees.  The space of $D$-matrices are also in bijection with the space of $F$-matrices.  However, the space of $F$-matrices has better separating properties using the $l^1$ or $l^2$ distances, as in Kim et al. (2020).
```{r}
dmat <- fmatrix:::Dmat_from_Fmat(mat)
dmat

fmatrix:::Fmat_from_Dmat(dmat)
```


## Nearby $F$-matrices

For a given matrix that is close to being an $F$-matrix but not quite one (such as the naive mean of a given sample of $F$-matrices), we might ask to see if there are any $F$-matrices close to it.  A heuristic is developed to compute a list of nearby $F$-matrices:

```{r}
not_mat

nearby_list <- nearby_Fmats(not_mat)

all(sapply(nearby_list, is.Fmat))

nearby_list
```
